---
title: "W207 Final Lab - EDA"
author: "Michael Winton"
date: "7/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(Hmisc)
```

## Load school data

```{r}
schools <- read.csv('2016 School Explorer.csv')
summary(schools[,1:41])  # non-grade-specific columns
# str(schools)
```

_Thoughts on columns to keep..._

Descriptive Columns:

- School.Name
- Location.Code (join to shsat$DBN)

Possible Explanatory Vars:

- District (integer)
- City (convert to factor)
- Zip (convert to factor)
- Grade.Low (e.g. PK; convert to factor)
- Grade.High (e.g. 12; convert to factor)
- Community.School (Yes/No; convert to factor)

- Economic.Need.Index (decimal)
- School.Income.Estimate (consider binning)

- Percent.ELL (convert to decimal)
- Percent.Asian  (convert to decimal)
- Percent.Black (convert to decimal)
- Percent.Hispanic (convert to decimal)
- Percent.Black...Hispanic (convert to decimal)
- Percent.White (convert to decimal)

- Student.Attendance.Rate (convert to decimal)
- Percent.of.Students.Chronically.Absent (convert to decimal)
- Rigorous.Instruction.. (convert to decimal)
- Collaborative.Teachers.. (convert to decimal)
- Supportive.Environment.. (convert to decimal)
- Effective.School.Leadership.. (convert to decimal)
- Strong.Family.Community.Ties.. (convert to decimal)
- Trust.. (convert to decimal)

- Average.ELA.Proficiency (decimal)
- Average.Math.Proficiency (decimal)

Comments:
- Common Core columns - we have options for percentages or buckets (both are provided)
- Propose ignoring the grade x demographic columns at least in initial model
- Dataet has one row per school

## Load SHSAT data

```{r}
shsat <- read.csv('D5 SHSAT Registrations and Testers.csv')
head(shsat)
str(shsat)
describe(shsat)
```

Comments:
- shsat$DBN = schools$Location.Code
- data has > 1 row per school
- consider aggregrating Year.of.SHST (ranges from 2013-2016)
- consider aggregating Grade.level (all are either 8 or 9)
- use Enrollment.on.10.31 to normalize the other columns
- also consider Enrollment.on.10.31 as an explanatory variable
- look at number registered as primary outcome var
- also look at % of registered who actually took the test as secondary potential outcome var

## Flatten `shsat` so we have one row per school (for joining)

Logic:
- aggregating 8th and 9th graders together
- creating separate columns for 2013, 2014, 2015, 2016
- do a full join, so even if some years had no registrations, the school will appear in the result set
- use DBN only; dropping school name (2 schools got renamed during this time period)

```{r}
# drop school name to avoid duplicate rows
tmp <- shsat %>% select(-School.name)

# aggregate Grade.level 8,9 together
tmp <- aggregate(. ~ DBN + Year.of.SHST, tmp, sum)

# drop Grade.level column now that we've aggregated
tmp <- tmp %>% select(-Grade.level)

# rename columns
# drop unnecessary ones
tmp2013 <- tmp %>%
  filter(Year.of.SHST==2013) %>%
  rename(enroll_2013 = Enrollment.on.10.31,
         num_reg_2013 = Number.of.students.who.registered.for.the.SHSAT,
         num_test_2013 = Number.of.students.who.took.the.SHSAT) %>%
  mutate(frac_reg_2013 = num_reg_2013/enroll_2013,
         frac_test_2013 = num_test_2013/num_reg_2013) %>%
  select(-Year.of.SHST)
tmp2014 <- tmp %>%
  filter(Year.of.SHST==2014) %>%
  rename(enroll_2014 = Enrollment.on.10.31,
         num_reg_2014 = Number.of.students.who.registered.for.the.SHSAT,
         num_test_2014 = Number.of.students.who.took.the.SHSAT) %>%
  mutate(frac_reg_2014 = num_reg_2014/enroll_2014,
         frac_test_2014 = num_test_2014/num_reg_2014) %>%
  select(-Year.of.SHST)
tmp2015 <- tmp %>%
  filter(Year.of.SHST==2015) %>%
  rename(enroll_2015 = Enrollment.on.10.31,
         num_reg_2015 = Number.of.students.who.registered.for.the.SHSAT,
         num_test_2015 = Number.of.students.who.took.the.SHSAT) %>%
  mutate(frac_reg_2015 = num_reg_2015/enroll_2015,
         frac_test_2015 = num_test_2015/num_reg_2015) %>%
  select(-Year.of.SHST)
tmp2016 <- tmp %>%
  filter(Year.of.SHST==2016) %>%
  rename(enroll_2016 = Enrollment.on.10.31,
         num_reg_2016 = Number.of.students.who.registered.for.the.SHSAT,
         num_test_2016 = Number.of.students.who.took.the.SHSAT) %>%
  mutate(frac_reg_2016 = num_reg_2016/enroll_2016,
         frac_test_2016 = num_test_2016/num_reg_2016) %>%
  select(-Year.of.SHST)

# join tables, resulting in one row per school
shsat_flat <- tmp2013 %>% full_join(tmp2014) %>% full_join(tmp2015) %>% full_join(tmp2016)

# review joined dataframe
head(shsat_flat)
summary(shsat_flat)
# describe(shsat_flat)
```

_NOTE: there are only 28 distinct schools in this dataset vs. 1272 in the first dataset.  So, as a first pass classification, we could simply classify as a "success" based on existence of data in the second dataset for the school (meaning they register at least some kids for the SHSAT test)._

Next, look at histograms of the numeric columns in this dataframe in case we want to further classify (e.g. high vs. low % registration).

### Plot histograms

```{r}
# fraction of students registered
par(mfrow=c(2,2))
hist(shsat_flat$frac_reg_2013, breaks=10)
hist(shsat_flat$frac_reg_2014, breaks=10)
hist(shsat_flat$frac_reg_2015, breaks=10)
hist(shsat_flat$frac_reg_2016, breaks=10)
```

Observations:
- Based on histograms, I'd consider classifying into 3 bins: Low (<0.3), Med(0.3-0.8), High(>0.8)

```{r}
# fraction of students who registered that actually take the test
par(mfrow=c(2,2))
hist(shsat_flat$frac_test_2013, breaks=10)
hist(shsat_flat$frac_test_2014, breaks=10)
hist(shsat_flat$frac_test_2015, breaks=10)
hist(shsat_flat$frac_test_2016, breaks=10)
```

Observations:
- Based on histograms, I'd consider classifying into 2 bins: Low (<0.6), High(>0.6)
- Getting kids to show up for the test doesn't appear to be quite as problematic as getting them to register
